### 2023-02-17 chapters/fairness/fif/experiments.tex
connecting chapters

\paragraph{Experimental Setup.} We implement a prototype of {\fairXplainer} in Python (version $ 3.7.6 $). To estimate FIFs, we leverage and modify the `HDMR' module in SALib library~\cite{Herman2017} based on global sensitivity analysis. In experiments, we consider four widely studied datasets from fairness literature, namely German-credit~\cite{DK2017},
% Ricci~\cite{mcginley2010ricci}, 
Titanic (\url{https://www.kaggle.com/c/titanic}), COMPAS~\cite{angwin2016machine}, and Adult~\cite{DK2017uci}. We deploy Scikit-learn~\cite{scikit-learn} to learn different classifiers: logistic regression classifier, support vector machine (SVM), neural network, and decision tree with $ 5 $-fold cross-validation. In experiments, we specify {\fairXplainer} to compute intersectional influences up to the second order ($ \lambda = 2 $). While applying cubic-spline based local regression in {\fairXplainer}, we set $ \tau $, the number of spline intervals to $ 6 $. We compare {\fairXplainer} with the existing Shapley-valued based FIF computational framework (\url{https://shorturl.at/iqtuX}), referred as SHAP~\cite{lundberg2020explaining}. For both {\fairXplainer} and SHAP, we set a timeout of $ 300 $ seconds for estimating FIFs. In addition, we deploy {\fairXplainer} along with a fairness-enhancing algorithm~\cite{kamiran2012data} and a fairness attack~\cite{solans2020poisoning} algorithm, and analyze the effect of these algorithms on the FIFs and the resultant fairness metric. In the following, we discuss the objectives of our empirical study.

>>>>

\paragraph{Experimental Setup.} We implement a prototype of {\fairXplainer} in Python (version $ 3.7.6 $). To estimate FIFs, we leverage and modify the `HDMR' module in SALib library~\cite{Herman2017} based on global sensitivity analysis. In experiments, we consider four widely studied datasets from fairness literature, namely German-credit~\cite{Dua:2019},
% Ricci~\cite{mcginley2010ricci}, 
Titanic (\url{https://www.kaggle.com/c/titanic}), COMPAS~\cite{angwin2016machine}, and Adult dataset~\cite{Dua:2019}. We deploy Scikit-learn~\cite{scikit-learn} to learn different classifiers: logistic regression classifier, support vector machine (SVM), neural network, and decision tree with $ 5 $-fold cross-validation. In experiments, we specify {\fairXplainer} to compute intersectional influences up to the second order ($ \lambda = 2 $). While applying cubic-spline based local regression in {\fairXplainer}, we set $ \tau $, the number of spline intervals to $ 6 $. We compare {\fairXplainer} with the existing Shapley-valued based FIF computational framework (\url{https://shorturl.at/iqtuX}), referred as SHAP~\cite{lundberg2020explaining}. For both {\fairXplainer} and SHAP, we set a timeout of $ 300 $ seconds for estimating FIFs. In addition, we deploy {\fairXplainer} along with a fairness-enhancing algorithm~\cite{kamiran2012data} and a fairness attack~\cite{solans2020poisoning} algorithm, and analyze the effect of these algorithms on the FIFs and the resultant fairness metric. In the following, we discuss the objectives of our empirical study.

### 2023-02-17 chapters/fairness/fvgm/introduction.tex
updated upto fvgm
connecting chapters

\label{chapter:fvgm}
	The significant improvement of machine learning over the decades has led to a host of applications of machine learning in  high-stake decision-making such as college admission~\cite{martinez2021using}, hiring of employees~\cite{ajunwa2016hiring}, and recidivism prediction~\cite{tollenaar2013method,dressel2018accuracy}. Machine learning algorithms often have an accuracy-centric learning objective, which may cause them to be biased towards certain part of the dataset belonging to a certain economically or socially sensitive groups~\cite{landy1978correlates,zliobaite2015relation,berk2019accuracy}.
%	Fairness in Machine Learning (ML) has gained a significant focus  in recent years as ML algorithms are deployed in high-stake decision-making, such as college admission, hiring of employees, and recidivism prediction. Thus, if the ML algorithm is biased towards certain part of the dataset belonging to certain economically or socially sensitive groups, it can cause adverse effects in real-life.
	 The following example illustrates a plausible case of unfairness of machine learning. 
	\begin{example}\label{fvgm_example:intro}
		\normalfont
		Following Example~\ref{fairness_justicia_example:intro}, let us consider a machine learning classification where the classifier decides the eligibility of an individual for health insurance given their income and fitness (Figure~\ref{fvgm_fig:example1}). Here, the sensitive feature `age' ($ A $) follows a Bernoulli distribution, and income ($ I $) and fitness ($ F $) follow Gaussian distributions. We generate 1000 samples from these distributions and use them to train a Support Vector Machine (SVM) classifier. The decision boundary of this classifier is $9.37I + 9.75F - 0.34A \ge 9.4$, where $A = 1$ denotes the sensitive group `age $ \ge $ 40'. This classifier selects an individual above and below $40$ years of age with probabilities $0.24$ and $0.86$, respectively. This illustrates a disparate treatment of individuals of two age groups by the SVM classifier. 
	\end{example}
%	The above example illustrates that even in simple scenarios, ML-based solutions can lead to disparate treatment. Motivated by studies indicating that such scenarios are rather not uncommon, there is a thrust in the research on the improvement of fairness in ML by proposing different notions of fairness definitions and fairness algorithms~\cite{hardt2016equality,kusner2017counterfactual,mehrabi2019survey}.

@@

	In order to solve the fairness-verification problem, existing works have proposed two principled approaches.	Firstly, \cite{ghosh2020justicia} and \cite{albarghouthi2017fairsquare} propose formal methods that reduce the problem into a solution of an SSAT or an SMT~ formula respectively.	Secondly, \cite{bastani2019probabilistic} propose a sampling approach that relies on extensively enumerating the conditional probabilities of prediction given different sensitive features and thus, incurs high computational cost. Additionally, existing works assume feature independence of non-sensitive features and consider correlated features within a limited scope, such as conditional probabilities of non-sensitive features w.r.t. sensitive features and ignore correlations among non-sensitive features. As a result, the \textit{scalability} and \textit{accuracy} of existing  verifiers remains a major challenge.

	In this work, we seek to remedy the aforementioned situation. As a first step, we focus on  \textit{linear classifiers}, which has attracted significant attention from researchers in the context of fair algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}. At this point, it is worth highlighting that our empirical evaluation demonstrates that the existing techniques fail to scale beyond small examples or provide highly inaccurate estimates for comparatively {\em small} linear classifiers.

%	In this paper, we argue that the above solution is not practical in verifying linear classifiers for two reasons: (i) the encoding of a linear classifier into an SSAT or an SMT formula or using a sampling approach suffer from scalability and (ii) ignoring correlations among all features results in an inaccurate estimate of fairness metrics.

%	Given multiple fairness definitions, fairness algorithms either \textit{preprocess} the training data~\cite{calmon2017optimized,kamiran2012data,zemel2013learning}, \textit{in-process} the classifier during training~\cite{zhang2018mitigating}, or \textit{post-process} the prediction of the trained classifiers~\cite{hardt2016equality,kamiran2012decision} to improve fairness. In contrast, fairness-attack algorithms try to manipulate the data or the classifier to worsen its fairness~\cite{solans2020poisoning}.

%	\red{Among different ML classifiers, researcher This observation motivates us in asking the following research question: \textit{Can we design an efficient fairness verification framework for linear classifiers that can verify different fairness definitions and algorithms accurately?}

%	
%	Considering the surge of research in ML fairness based on linear classifiers, we focus on designing a fairness verifier particularly tailored for linear classifiers. We further show novel applications of the verifier in practical fairness tasks, such as detecting fairness attacks and computing the source of unfairness in the level of input features.
%

	\textbf{Our Contributions.} The contributions of this chapter are summarized below.

>>>>

\label{chapter:fvgm}
\begin{comment}
	The significant improvement of machine learning over the decades has led to a host of applications of machine learning in  high-stake decision-making such as college admission~\cite{martinez2021using}, hiring of employees~\cite{ajunwa2016hiring}, and recidivism prediction~\cite{tollenaar2013method,dressel2018accuracy}. Machine learning algorithms often have an accuracy-centric learning objective, which may cause them to be biased towards certain part of the dataset belonging to a certain economically or socially sensitive groups~\cite{landy1978correlates,zliobaite2015relation,berk2019accuracy}.
%	Fairness in Machine Learning (ML) has gained a significant focus  in recent years as ML algorithms are deployed in high-stake decision-making, such as college admission, hiring of employees, and recidivism prediction. Thus, if the ML algorithm is biased towards certain part of the dataset belonging to certain economically or socially sensitive groups, it can cause adverse effects in real-life.
	 The following example illustrates a plausible case of unfairness of machine learning. 
	\begin{example}\label{fvgm_example:intro}
		\normalfont
		Following Example~\ref{fairness_justicia_example:intro}, let us consider a machine learning classification where the classifier decides the eligibility of an individual for health insurance given their income and fitness (Figure~\ref{fvgm_fig:example1}). Here, the sensitive feature `age' ($ A $) follows a Bernoulli distribution, and income ($ I $) and fitness ($ F $) follow Gaussian distributions. We generate 1000 samples from these distributions and use them to train a Support Vector Machine (SVM) classifier. The decision boundary of this classifier is $9.37I + 9.75F - 0.34A \ge 9.4$, where $A = 1$ denotes the sensitive group `age $ \ge $ 40'. This classifier selects an individual above and below $40$ years of age with probabilities $0.24$ and $0.86$, respectively. This illustrates a disparate treatment of individuals of two age groups by the SVM classifier. 
	\end{example}
%	The above example illustrates that even in simple scenarios, ML-based solutions can lead to disparate treatment. Motivated by studies indicating that such scenarios are rather not uncommon, there is a thrust in the research on the improvement of fairness in ML by proposing different notions of fairness definitions and fairness algorithms~\cite{hardt2016equality,kusner2017counterfactual,mehrabi2019survey}.

@@

\end{comment}

	In this chapter, we extend the probabilistic fairness verification problem to a more practical scenario by accounting for the correlation of features, which is prevalent in most classification datasets. Existing fairness verifiers \cite{albarghouthi2017fairsquare,bastani2019probabilistic}, including Justicia \cite{ghosh2020justicia} in Chapter~\ref{chapter:justicia}, suffer from poor accuracy due to considering specific input distribution. For example, they assume feature independence of non-sensitive features and consider correlated features within a limited scope, such as conditional probabilities of non-sensitive features with respect to sensitive features, and ignore correlations among non-sensitive features. Thus, it is desirable to design a fairness verification framework where the input is a probability distribution containing feature correlations.

	Towards this goal, we particularly focus on the verification of \textit{linear classifiers} as our first step because there has been significant attention on linear classifiers in the context of fair algorithms \cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}. In this context, existing approaches suffer from limited scalability while verifying linear classifiers. This is due to the encoding of linear classifiers into SSAT/SMT formulas \cite{ghosh2020justicia,albarghouthi2017fairsquare}. For example, Justicia applies pseudo-Boolean to CNF translation of linear classifiers as a pre-processing step, and the encoding becomes large depending on the number of features and the precision of real-valued coefficients in linear classifiers.

	Considering these two aspects, our goal in this chapter is to design a fairness verifier particularly tailored for linear classifiers that addresses both the scalability and accuracy challenges of existing verifiers.

	\paragraph{Our Contributions.} The contributions of this chapter are summarized below.

### 2023-02-17 chapters/fairness/justicia/experiments.tex
connecting chapters

For comparative evaluation of {\justicia}, we have experimented with two state-of-the-art probabilistic fairness verifiers FairSquare and VeriFair, and a sample-based fairness measuring tool: AIF360. In the experiments, we have studied three type of classifiers: decision tree, logistic regression classifier, and CNF learner. Decision tree and logistic regression are implemented using scikit-learn module of Python~\cite{PVGMTGB2011} and we use the MaxSAT-based CNF learner, namely IMLI~\cite{GM2019}. We have used the PySAT library~\cite{imms-sat18} for encoding the decision function of the logistic regression classifier into a CNF formula. In our experiments, we have verified two fairness-enhancing algorithms: reweighing algorithm~\cite{kamiran2012data} and optimized pre-processing  algorithm~\cite{calmon2017optimized}. 
We have experimented on multiple datasets containing multiple sensitive features: the UCI\footnote{\url{ http://archive.ics.uci.edu/ml}} Adult and German-credit dataset,  ProPublica’s COMPAS recidivism dataset~\cite{angwin2016machine}, Ricci dataset~\cite{mcginley2010ricci}, and Titanic dataset\footnote{\url{https://www.kaggle.com/c/titanic}}.
\iffalse 
Since both {\justicia}  and FairSquare take a  probability distribution of the features as input, we perform five-fold cross validation, use the train set for learning the classifier, compute distribution on the test set and finally verify fairness metrics such as disparate impact and statistical parity difference on the distribution. 
\fi
%that pre-process the dataset to mitigate its bias

@@

\subsubsection{Robustness: Stability to Sample Size} 
We have compared the robustness of our probabilistic fairness verifier {\justicia} with dataset-centric verifier AIF360 by varying the sample-size and reporting the standard deviation of different fairness metrics. 
In Figure~\ref{fairness_justicia_fig:sample-size}, AIF360 shows higher standard deviation for lower sample-size and the value decreases as  the sample-size increases. 
In contrast, {\justicia} shows significantly lower ($\sim10\times$ to $100\times$) standard deviation for different sample-sizes. 
The reason is that AIF360 empirically measures on a fixed dataset whereas {\justicia} provides estimates over the distribution.
Thus, {\justicia} is more robust than the sample-based verifier AIF360.

>>>>

For comparative evaluation of {\justicia}, we have experimented with two state-of-the-art probabilistic fairness verifiers FairSquare and VeriFair, and a sample-based fairness measuring tool: AIF360. In the experiments, we have studied three type of classifiers: decision tree, logistic regression classifier, and CNF learner. Decision tree and logistic regression are implemented using scikit-learn module of Python~\cite{scikit-learn} and we use the MaxSAT-based CNF learner, namely IMLI~\cite{GM2019}. We have used the PySAT library~\cite{imms-sat18} for encoding the decision function of the logistic regression classifier into a CNF formula. In our experiments, we have verified two fairness-enhancing algorithms: reweighing algorithm~\cite{kamiran2012data} and optimized pre-processing  algorithm~\cite{calmon2017optimized}. 
We have experimented on multiple datasets containing multiple sensitive features: the UCI\footnote{\url{ http://archive.ics.uci.edu/ml}} Adult and German-credit dataset,  ProPublica’s COMPAS recidivism dataset~\cite{angwin2016machine}, Ricci dataset~\cite{mcginley2010ricci}, and Titanic dataset\footnote{\url{https://www.kaggle.com/c/titanic}}.
\iffalse 
Since both {\justicia}  and FairSquare take a  probability distribution of the features as input, we perform five-fold cross validation, use the train set for learning the classifier, compute distribution on the test set and finally verify fairness metrics such as disparate impact and statistical parity difference on the distribution. 
\fi
%that pre-process the dataset to mitigate its bias

@@

\subsubsection{Robustness: Stability to Sample Size} 
We have empirically compared the robustness of our probabilistic fairness verifier {\justicia} with dataset-centric verifier AIF360 by varying the sample-size and reporting the standard deviation of different fairness metrics. 
In Figure~\ref{fairness_justicia_fig:sample-size}, AIF360 shows higher standard deviation for lower sample-size and the value decreases as  the sample-size increases. 
In contrast, {\justicia} shows significantly lower ($\sim10\times$ to $100\times$) standard deviation for different sample-sizes. 
The reason is that AIF360 empirically measures on a fixed dataset whereas {\justicia} provides estimates over the distribution.
Thus, {\justicia} is more robust than the sample-based verifier AIF360.

### 2023-02-17 chapters/fairness/justicia/introduction.tex
connecting chapters

\paragraph{Contribution.} From this vantage point, \textit{we propose to model verifying different fairness metrics as a Stochastic Boolean Satisfiability (SSAT) problem}~\cite{littman2001stochastic}. SSAT was originally introduced by ~\cite{papadimitriou1985games} to model {\em games against nature}. In this work, we primarily focus on reductions to the exist-random quantified fragment of SSAT, which is also known as E-MAJSAT~\cite{littman2001stochastic}.   SSAT is a conceptual framework that has been employed to capture several fundamental problems in artificial intelligence such as the computation of maximum a posteriori (MAP) hypothesis~\cite{fremont2017maximum},  propositional probabilistic planning~\cite{majercik2007appssat},  and circuit verification~\cite{lee2018towards}. Furthermore, our choice of SSAT as a target formulation is motivated by the recent algorithmic progress that has yielded efficient SSAT tools~\cite{lee2017solving,lee2018solving}.

>>>>

\paragraph{Contribution.} From this vantage point, \textit{we propose to model verifying different fairness metrics as a Stochastic Boolean Satisfiability (SSAT) problem}~\cite{littman2001stochastic}. SSAT was originally introduced by ~\cite{papadimitriou1985games} to model {\em games against nature}. In this chapter, we primarily focus on reductions to the exist-random quantified fragment of SSAT, which is also known as E-MAJSAT~\cite{littman2001stochastic}.   SSAT is a conceptual framework that has been employed to capture several fundamental problems in artificial intelligence such as the computation of maximum a posteriori (MAP) hypothesis~\cite{fremont2017maximum},  propositional probabilistic planning~\cite{majercik2007appssat},  and circuit verification~\cite{lee2018towards}. Furthermore, our choice of SSAT as a target formulation is motivated by the recent algorithmic progress that has yielded efficient SSAT tools~\cite{lee2017solving,lee2018solving}.

### 2023-02-17 chapters/fairness/justicia/practical.tex
connecting chapters

\subsubsection{Extension to Continuous Features}
In practical problems, features are generally real-valued or categorical but classifiers, which are naturally expressed as CNF such as~\cite{GMM20}, are generally trained on a Boolean abstraction of input features. In order to perform the Boolean abstraction, each categorical feature is one-hot encoded and each real-valued feature is discretized into a set of Boolean features~\cite{LKCL2019,GMM20}.

>>>>

\subsubsection{Extension to Continuous Features}
In practical problems, features are generally real-valued or categorical but classifiers, which are naturally expressed as CNF such as~\cite{GMM2020}, are generally trained on a Boolean abstraction of input features. In order to perform the Boolean abstraction, each categorical feature is one-hot encoded and each real-valued feature is discretized into a set of Boolean features~\cite{LKCL2019,GMM2020}.

### 2023-02-17 chapters/interpretability/imli/introduction.tex
connecting chapters

%\section{Introduction}

%The last decade has witnessed significant progress in machine learning (ML). 
\label{chapter:imli}
Machine learning models are presently deployed in safety-critical and high-stake decision-making arising in medical~\cite{erickson2017machine,kaissis2020secure,kononenko2001machine}, law~\cite{kumar2018law,surden2014machine}, and transportation~\cite{peled2019model,zantalis2019review}. In safety-critical domains, a surge of interest has been observed in designing interpretable, robust, and fair models instead of merely focusing on an accuracy-centric learning objective~\cite{bhagoji2018enhancing,doshi2017towards,du2019techniques,hancox2020robustness,holstein2019improving,mehrabi2021survey,molnar2020interpretable,murdoch2019interpretable}. In this chapter, we focus on learning an interpretable machine learning classifier~\cite{rudin2019stop}, where the rules governing the prediction are explicit (by design) in contrast to black-box models.

%Machine learning techniques are now being deployed in critical decision making, for example, hiring employees, giving a loan to people, predicting recidivism, predicting the likelihood of stroke of patients, etc. In the classical setting of classification problems, making an accurate prediction has been the objective of several prediction systems e.g., ad predictions, movie recommendations, etc. at the cost of making such systems essentially black-box. However, there are other safety-critical domains e.g., legal, medical, law, and so on, where interpretability, privacy, and fairness of the prediction system are of paramount importance. In this study, we show a principled approach to generate accurate and interpretable decision functions expressible in propositional logic.

In order to achieve the interpretability of predictions, decision functions in the form of classification rules such as decision trees,  decision lists, decision sets, etc.\ are particularly effective~\cite{bessiere2009minimising,dash2021lprules,ignatiev2021reasoning,izza2020explaining,lakkaraju2017interpretable,lakkaraju2016interpretable,letham2015interpretable,narodytska2018learning,rivest1987learning,wang2015falling,yu2020optimal}. Such models are used to either learn an interpretable model from the start or as proxies to provide post-hoc explanations of pre-trained black-box models~\cite{gill2020responsible,lundberg2017unified,moradi2021post,ribeiro2016should,slack2020fooling}.  In our context of rule-based classifiers, we use the sparsity of rules as a measure of interpretability, which is adopted in various domains, specifically in the medical domain~\cite{gage2001validation,lakkaraju2019faithful,letham2015interpretable,malioutov2013exact,myers1962myers}.

In this chapter, we study decision functions expressible in \emph{propositional logic}. In particular, we focus on learning logical formulas from data as classifiers and define interpretability in terms of the number of Boolean literals that the formula contains. In the following, we illustrate an interpretable classifier that decides if a tumor cell is malignant or benign based on different features of tumors.

@@

\begin{example}
	\normalfont
	We learn an interpretable classification rule in CNF for predicting the classification of a tumor cell. We consider WDBC (Wisconsin Diagnostic Breast Cancer) dataset~\cite{agarap2018breast} to learn a CNF formula with two \emph{clauses} and four Boolean \emph{literals}. In the formula, clauses are connected by Boolean `AND', where each clause contains \emph{literals} connected by Boolean `OR'. Informally, a tumor cell is malignant if at least one literal in each clause becomes true.
\end{example}

>>>>

\label{chapter:imli}
%Machine learning models are presently deployed in safety-critical and high-stake decision-making arising in medical~\cite{erickson2017machine,kaissis2020secure,kononenko2001machine}, law~\cite{kumar2018law,surden2014machine}, and transportation~\cite{peled2019model,zantalis2019review}. In safety-critical domains, a surge of interest has been observed in designing interpretable, robust, and fair models instead of merely focusing on an accuracy-centric learning objective~\cite{bhagoji2018enhancing,doshi2017towards,du2019techniques,hancox2020robustness,holstein2019improving,mehrabi2021survey,molnar2020interpretable,murdoch2019interpretable}. 
In this chapter, we learn an interpretable rule-based classifier~\cite{rudin2019stop}, where the rules governing the prediction are explicit (by design) in contrast to black-box classifiers. 
%In order to achieve the interpretability of predictions, decision functions in the form of classification rules such as decision trees,  decision lists, decision sets, etc.\ are particularly effective~\cite{bessiere2009minimising,dash2021lprules,ignatiev2021reasoning,izza2020explaining,lakkaraju2017interpretable,lakkaraju2016interpretable,letham2015interpretable,narodytska2018learning,rivest1987learning,wang2015falling,yu2020optimal}. Such models are used to either learn an interpretable model from the start or as proxies to provide post-hoc explanations of pre-trained black-box models~\cite{gill2020responsible,lundberg2017unified,moradi2021post,ribeiro2016should,slack2020fooling}.  In our context of rule-based classifiers, we use the sparsity of rules as a measure of interpretability, which is adopted in various domains, specifically in the medical domain~\cite{gage2001validation,lakkaraju2019faithful,letham2015interpretable,malioutov2013exact,myers1962myers}.  
In particular, we study classifiers expressed as Boolean formulas, wherein we define interpretability in terms of the number of Boolean literals that the formula contains. Here, we illustrate an interpretable classifier that decides if a tumor cell is malignant or benign based on different features of tumors.

@@

\begin{example}
	We illustrate an interpretable classification rule in CNF for classifying a tumor cell. We consider WDBC (Wisconsin Diagnostic Breast Cancer) dataset~\cite{agarap2018breast} to learn a CNF formula with two \emph{clauses} and four Boolean \emph{literals}. In the formula, clauses are connected by Boolean `AND', where each clause contains \emph{literals} connected by Boolean `OR'. Informally, a tumor cell is malignant if at least one literal in each clause becomes true.
\end{example}
